Think deep for reef resilience

### 2025-09-02 paper/paper_draft.qmd
Various paper updates
Updated:
- abstract to show interaction with incoming connectivity
- exploratory modelling references
- supplementary figure showing the interaction between depth and outgoing connectivity
- Add initial discussion paragraph outlining high level findings
- Move GCM discussion to later in the flow (need to check if it still flows ok)
- Added discussion around outgoing connectivity and the potential for managers to target reefs with high depth, incoming connectivity and outgoing connectivity
- Added limitation caveat: We did not include depth effects on growth due to decreas

---
title: Think deep for reef resilience
author:
  - name: Benjamin Grier
    email: grierben777@gmail.com
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          state: QLD
    attributes:
        corresponding: true
  - name: Takuya Iwanaga
    email: t.iwanaga@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
    attributes:
        corresponding: true
abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great Barrier
    Reef (GBR) experiencing cumulative stressors and increasingly severe declines in coral cover from thermal stress events.
    Understanding drivers behind reef resilience to climate impacts is critical for conservation
    planning and intervention strategies. A scenario-based population modelling approach was
    adopted with larval connectivity dynamics, thermal adaptation, and environmental factors
    to assess coral reef resilience across GBR locations under future climate projections.
    Using five Global Climate Models (GCMs) spanning five Equilibrium Climate Sensitivities,
    we analysed reef coral cover time series and the ability of reefs to maintain positive carbonate
    production budgets under future conditions. Contrary to expectations, larval connectivity
    patterns did not correlate with increases in maintenance of positive carbonate budgets.
    Instead, reef depth emerged as the primary predictor of cover-time series resilience and
    maintenance of positive carbonate budgets, with deeper reefs benefiting from thermal
    refugia that reduce bleaching impacts. Interaction between reef depth and incoming 
    connectivity was found to influence reef temporal behaviour. Our findings suggest that 
    depth provides critical protection against thermal stress, contrasting current restoration
    practices that primarily target easily accessible shallow sites. These results have 
    important implications for reef management, indicating that depth refugia should be 
    prioritized for further consideration in conservation planning alongside connectivity 
    considerations to maximize coral survival under continuing climate change.
keywords:
  - Coral reef
  - Climate change
  - Ecosystem resilience
  - Reef depth
  - Reef connectivity
date: last-modified
format:
  elsevier-docx:
    journal:
      name: Journal Name
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
engine: julia
crossref:
  custom:
  - kind: float
    reference-prefix: SFigure
    key: sfig
  custom:
  - kind: float
    reference-prefix: STable
    key: stbl
editor:
  markdown:
    wrap: 92
    canonical: true
---

@@

Climate change is impacting species survival, persistence and ecosystem functions globally. This is particularly true for coral reefs, which are vulnerable to increasing magnitude and variation in environmental stressors such as ocean temperature, acidification and nutrient loading [@foster2021; @gattuso2014]. Mass bleaching and mortality events driven by climate change can cause reef-building coral population declines, possibly leading to long-term and large-scale coral-macroalgae regime shifts [@arif2022; @wild2011]. Climate impacts on coral reefs are decreasing the ability of coral reefs to provide ecological functions such as benthic habitat and nutrient cycling [@forster2021; @williams2019] as well as ecosystem services such as food, protection, tourism and Indigenous cultural value [@eddy2021; @stoeckl2021].

Understanding the characteristics of reefs that will persist under projected climate conditions requires examining reef resilience across multiple scales and processes. Structural resilience - the capacity to maintain coral populations and avoid ecosystem regime shifts - depends on three interconnected mechanisms. First, the ability of individual corals to survive thermal stress. Second, the connectivity networks that enable recolonisation after disturbances. Third, the maintenance of the carbonate framework that provides reef structure. We aim to assess how these mechanisms interact under projected climate conditions to identify the characteristics of structurally resilient reefs through scenario-based coral population modelling for the Great Barrier Reef ecosystem.

The Great Barrier Reef (GBR) is the world's largest coral reef ecosystem consisting of thousands of individual reefs. They are interconnected through coral and fish larval dispersal patterns mediated by hydrodynamic, spatial, and ecological processes [@grimaldi2022]. The GBR faces severe projected coral cover declines under climate change [@bozec2025; @bozec2022]. These coral cover declines will be compounded by ocean acidification, tropical cyclones, and crown-of-thorns starfish outbreaks [@death2012; @fox-kemper2021; @seneviratne2021; @sweatman2011; @wolff2018]. The 2023-24 Summer mass bleaching event may be signalling a shift towards this reef future, with some level of bleaching recorded on 79% of surveyed reefs (total 1001 surveyed reefs) [@cantin2024].

@@

An exploratory scenario modelling approach was adopted for this investigation. The method involves exploring a wide range of plausible futures under different assumptions of environmental (climatic) conditions and ecological factors. Exploratory scenario modelling embraces uncertainty as an inherent characteristic of complex environmental systems [@lempert2010] and is a core methodology underpinning the development of robust and adaptive management strategies [@haasnoot2013].

Exploratory modelling approaches have gained favour in recent years as precise long-term predictions of future conditions are often impossible in complex systems like coral reefs. Multiple interacting and compounding sources of uncertainty, and a lack of broad agreement as to their likelihood, create inherently uncertain outcomes. Rather than attempting to forecast specific outcomes, exploratory modelling aims to systematically explore many possible futures to identify patterns. Allowing for the identification of robust management strategies that perform well across those diverse conditions, as well as conditions in which such strategies may fail. Such an exploratory approach lends itself well to the purpose of the investigation. Specifically, to assess the role of thermal stress, connectivity, and depth to identify the environmental and reef characteristics that support structural resilience.

@@

To apply exploratory scenario modelling, we employed a modelling framework involving two separate systems: ADRIA (the Adaptive Dynamic Reef Intervention Algorithms platform) and CoralBlox. ADRIA is a multi-criteria decision support platform designed to help reef managers address climate change challenges in coral reef systems [@iwanaga2023]. The ADRIA platform provides functionality for scenario generation and analysis as well as providing the technical infrastructure to couple environmental and ecosystem representations. CoralBlox is a computationally efficient coral population model. It simulates functional/morphological groups of corals over their life cycles, competing for available space over time [@ribeirodealmeida2024]. The coupled framework is referred to as ADRIA-CoralBlox.

@@

Reefs in each time series cluster were compared to identify characteristics that differ between Low, Medium and High clusters. These characteristics include reef median depth, mean time series DHW, Pearson’s correlation between total coral cover and inverse DHW time series, proportion of coral habitable area occupied by corals in the first timestep, ratio of source-to-sink larval connectivity and weighted incoming connectivity.

@@

In many bioregions reefs in the Low cluster have the lowest correlation between coral cover and inverse-DHW time series whereas reefs in the High cluster were found to have the strongest correlation (@sfig-access-esm1-5-bioregion-dhwcor). Across six of the 29 bioregions reef mean DHW levels increased as cluster cover decreased (@sfig-access-esm1-5-bioregion-dhwmean). There were no consistent patterns across bioregions in their reefs' initial proportion of habitable area occupied by corals and cluster assignment (@sfig-access-esm1-5-bioregion-initialproportion). Across most bioregions there were not consistent differences between clusters in connectivity metrics such as Log-10 source-to-sink connectivity ratio (@sfig-access-esm1-5-bioregion-logsosi) and Log-10 weighted incoming connectivity (@sfig-access-esm1-5-bioregion-weightedincoming).

@@

Analyses for source-to-sink ratio and weighted incoming and outgoing connectivity metrics suggest that reefs with greater outgoing connectivity and lower incoming connectivity levels have higher probability of being clustered into the Low cluster category compared to other reefs. Additionally, two-way analyses revealed an interaction between reef depth and incoming connectivity in determining cluster category assignment, particularly at the high depth range (@fig-ts-cluster-pdp-twoway). Two-way analyses revealed that reefs at greater than 10m depth are unlikely to be clustered into the Low cluster category, regardless of their incoming connectivity levels. However, at greater than 10m depth reefs with higher incoming connectivity are more likely to be assigned to a High cluster, while reefs with low incoming connectivity are more likely to be assigned to a Medium cover cluster (@fig-ts-cluster-pdp-twoway).

Individual feature analysis for carrying capacity indicates that its influence on cluster assignment is concentrated at the low range of predictor values (@fig-ts-cluster-pdp). This aligns with the right-skewed distribution of carrying capacity across reefs for these three predictor variables. Beyond this low-end range, the partial dependence becomes near constant, contrary to their feature importance scores which suggest these characteristics do contribute important information to predictions. The contrasting outcomes indicate that interactions between other reef characteristics are driving behaviour rather than a strong influence from a single individual characteristic.

Assessing random forest prediction accuracy across predictor values, summarised by bioregions, reveals increasing performance as bioregion mean reef depth increases (@sfig-rf-predictor-perf). This highlights that there may be unexplained variation between reefs in shallow bioregions that contributes to cluster assignment and reef cover outcomes but is not captured by our included predictor variables.

@@

![Partial dependence plot indicating the probability of cluster assignment along interacting values of median reef depth (metres) and Log10 weighted incoming connectivity. Colour indicates the probability of assignment to each cluster, with lighter colours indicating higher probability of assignment. Panel A shows the probability of assignment for the Low cover cluster, Panel B shows Medium cover cluster and Panel C shows assignment probability for the High cluster category. Probability values for each cluster are generated by the random forest classifier using each combination of depth and incoming connectivity values along the axes at 50 grid points. The actual distributions of depth and incoming connectivity values for reefs are displayed in panels D and E respectively.](../figs/ts_cluster_rf_twoway_depth_conn_pdp.png){#fig-ts-cluster-pdp-twoway fig-align="center" width="14.82cm"}

@@

Reef median depth had a moderate positive correlation with the number of years a reef maintains coral cover above the assumed carbonate budget threshold required to sustain live coral cover ($\theta_{0}$). Greater median depth was a key characteristic defining reefs that sustain coral cover above $\theta_{0}$ (@fig-access-esm1-5-depth-conn-carbonate; @sfig-access-esm1-5-mean-carb-budget). Unsurprisingly, the number of years above $\theta_0$ substantially decreased as $\theta_0$ is increased, as does the Spearman's correlation. Decreasing correlation levels are likely due to an increasing number of reefs being unable to maintain coral cover above $\theta_0$ as it is increased, with the distribution of years above threshold becoming increasingly right-skewed (@fig-access-esm1-5-depth-conn-carbonate). In contrast, log10-transformed total connectivity strength ($\text{log10}(S)$) was found to have no correlation with the number of years a reef maintains cover above the required thresholds (@fig-access-esm1-5-depth-conn-carbonate). The correlation between $\text{log10}(S)$ and $\theta_0$ slightly increased as the assumed threshold for $\theta_0$ increased.

![Number of years reef coral cover exceeds the carbonate budget threshold from 2025 to 2075 (y-axis). The assessed carbonate budget thresholds range from 10 to 20% of total reef area. Bracketed x-axis values represent the Spearman's correlation between reef variables (depth - left and Log-10 connectivity strength - right) and the number of years cover exceeds the carbonate budget threshold. Point colour represents the median reef depth in metres in the left-hand panel, and Log-10 total connectivity strength in the right-hand panel. Red points represent the mean number of years above an assumed carbonate budget threshold across reefs. Deeper reefs are better able to maintain cover above the indicative carbonate budget threshold, while there is no considerable relationship between connectivity strength and carbonate budget maintenance. Displays results from median time series using ACCESS-ESM1-5.](../figs/ACCESS-ESM1-5/depth_conn_carbonate_comparison.png){#fig-access-esm1-5-depth-conn-carbonate fig-align="center" width="14.82cm"}

@@

@@

Reefs in different clusters largely do not differ in mean modelled DHW levels experienced and exhibit overall similar DHW time series. While these reefs experience similar magnitude DHWs based on SST, the proportion of the reefs' coral population impacted by heat stress events depends on reef depth. Reefs at deeper depths are likely to be clustered into High cover clusters. Reefs at shallow depths experience no, or few, years with coral cover exceeding levels likely required to maintain positive carbonate budgets. Movement of shallow reefs towards negative carbonate budgets in the future may contribute to phase shifts in reef dominant CaCO~3~ production. Shifting from coral driven to encrusting algae driven production, or to bioerosion dominated systems with rapid erosion of physical reef frameworks [@dudgeon2010; @perry2008]. While reefs at increased depths may be able to maintain coral driven CaCO~3~ production and accumulation into the future for longer. Our results suggest that within shallow-water reef ecosystems (non-mesophotic, depth \< 30m), depth contributes greatly to reef resilience and the ability of the reef to maintain positive carbonate budgets under future conditions. While similar findings have been demonstrated in the Pacific [@smith2014; @baird2018] and Indo-Pacific [@bridge2014] regions, depth refugia analyses have often focussed on the potential of mesophotic reefs (\> 30m deep) as thermal refugia [@bongaerts2010; @holstein2015; @semmler2017]. In contrast, Venegas et al. [-@venegas2019] found no association between depth and the alleviation of heat stress across a range of shallow reefs in the Pacific.

These findings support the hypothesis that reefs at greater depths will exhibit characteristics of refugia and more prolonged structural resilience under climate change compared to shallower reefs. Additionally, Liu, Callaghan and Baldock [-@liu2024] suggest that rubble instability is inversely correlated with reef depth, possibly limiting juvenile recruitment capacity of shallow reefs. A caveat to these depth-resilience findings is the quality of the bathymetry data used to calculate median reef depth. As detailed previously, estimates of median reef depth are derived from satellite imagery (at $10m^2$ resolution) which can be subject to high degrees of uncertainty, both in terms of the estimated depth at for each pixel as well as their alignment to the established GBRMPA reef outlines.

Current reef restoration and intervention-investigation efforts on the GBR, such as deployment of coral seeding and fragment devices, target reefs with suitable restoration sites in the 1-10m depth range [@howlett2021; @randall2021; @whitman2024]. While there are no explicit justifications for this depth range, restoration depth is likely constrained by snorkelling and diving deployment and monitoring logistical accessibility, visual identification of sites from the surface, in combination with tourism return value. Our results suggest that of the assessed reefs occurring within the 1-10m depth range (1736 reefs), 57% (991) of reefs fall into the low and medium clusters (across all GCMs at the GBRMPA management region scale), with many unable to maintain positive carbonate budgets under future conditions. Partial dependence analysis revealed a range of depths from 10-15m where the probability of a reef being clustered into the High cover cluster increased substatially. Additionally, two-way partial dependence analysis revealed that high incoming connectivity is important for High cluster assignment for reefs at depths of \>10m. Considering this 10-15m depth range, and targetting well connected reefs, when assessing possible restoration sites may reduce the exposure of planted and seeded corals to high thermal stress, while still maintaining effective logistical access to sites. Although transient, and difficult to measure, properties such as reef connectivity are likely to increase uncertainty in intervention planning to a greater degree than environmental characteristics such as reef depth. Reef intervention planning is a complex and challenging decision-making process [@anthony2020], and our results suggest that deployment site depth may be an important factor to additionally consider when assessing the long-term resilience of sites.

@@

There are multiple caveats in how coral ecosystem dynamics are modelled in this investigation. We model coral colonies as circular diameter values that are tracked over timesteps through growth. Additionally, we do not model the direct influence of light intensity on coral growth or the negative influence of depth on coral growth rates due to reduced light levels. \[**BG: Include citations for influence of depth/light on growth rates**\]. If included in this investigation, it is possible that reef recovery ability would be reduced at greater depths even though corals at these depths experience refugia from thermal stress. Due to the availability of quality future projected data we do not consider other important impacts such as waves, cyclones, currents and Crown of Thorns Starfish (CoTS) on coral colony survival and bleaching [@massel1993; @deaker2022].

@@

@@

@@

![Total coral cover – DHW Pearson’s correlation levels for reefs within each bioregion, separated into clusters assigned by time series clustering. ‘Low’ clusters (green) are clusters with the lowest median area-relative cover between 2030 and 2060 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median area-relative cover between 2030 and 2060 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median area-relative cover between 2030 and 2060 within each bioregion. Solid lines represent the median cover – DHW correlation level for each cluster and bands represent the distribution of correlation levels within each cluster. Displays results for ACCESS-ESM1-5 GCM.](../figs/ACCESS-ESM1-5/bioregion/dhw_cover_cor_bioregion_violin.png){#sfig-access-esm1-5-bioregion-dhwcor fig-align="center" width="14.82cm"}

>>>>

---
title: Think deep for reef resilience
author:
  - name: Benjamin Grier
    email: grierben777@gmail.com
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          state: QLD
    attributes:
        corresponding: true
  - name: Takuya Iwanaga
    email: t.iwanaga@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
    attributes:
        corresponding: true
abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great Barrier
    Reef (GBR) experiencing cumulative stressors and increasingly severe declines in coral cover from thermal stress events.
    Understanding drivers behind reef resilience to climate impacts is critical for conservation
    planning and intervention strategies. A scenario-based population modelling approach was
    adopted with larval connectivity dynamics, thermal adaptation, and environmental factors
    to assess coral reef resilience across GBR locations under future climate projections.
    Using five Global Climate Models (GCMs) spanning five Equilibrium Climate Sensitivities,
    we analysed reef coral cover time series and the ability of reefs to maintain positive carbonate
    production budgets under future conditions. Contrary to expectations, larval connectivity
    patterns did not correlate with increases in maintenance of positive carbonate budgets.
    Instead, reef depth emerged as the primary predictor of cover-time series resilience and
    maintenance of positive carbonate budgets, with deeper reefs benefiting from thermal
    refugia that reduce bleaching impacts. Interaction between reef depth and incoming 
    connectivity was found to influence reef temporal behaviour. Our findings suggest that 
    depth provides critical protection against thermal stress, with high incoming connectivity enhancing the performance of deep reefs, contrasting current restoration
    practices that primarily target easily accessible sites in the 1-10m depth range. These results have 
    important implications for reef management, indicating that depth refugia should be 
    prioritized for further consideration in conservation planning alongside connectivity 
    considerations to maximize coral survival under continuing climate change.
keywords:
  - Coral reef
  - Climate change
  - Ecosystem resilience
  - Reef depth
  - Reef connectivity
date: last-modified
format:
  elsevier-docx:
    journal:
      name: Journal Name
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
engine: julia
crossref:
  custom:
  - kind: float
    reference-prefix: SFigure
    key: sfig
  custom:
  - kind: float
    reference-prefix: STable
    key: stbl
editor:
  markdown:
    wrap: 92
    canonical: true
---

@@

Climate change is impacting species survival, persistence and ecosystem functions globally. This is particularly true for coral reefs, which are vulnerable to increasing magnitude and variation in environmental stressors such as ocean temperature, acidification and nutrient loading [@foster2021; @gattuso2014]. Mass bleaching and mortality events driven by climate change can cause reef-building coral population declines [@hughes2017; @hughes2018; @bozec2025], possibly leading to long-term and large-scale coral-macroalgae regime shifts [@arif2022; @wild2011]. Climate impacts on coral reefs are decreasing the ability of coral reefs to provide ecological functions such as benthic habitat and nutrient cycling [@forster2021; @williams2019] as well as ecosystem services such as food, protection, tourism and Indigenous cultural value [@eddy2021; @stoeckl2021].

Understanding the characteristics of reefs that will persist under projected climate conditions requires examining reef resilience across multiple scales and processes. Structural resilience - the capacity to maintain coral populations and avoid ecosystem regime shifts - depends on three interconnected mechanisms. First, the ability of individual corals to survive thermal stress. Second, the connectivity networks that enable recolonisation after disturbances [@hock2017; @mumby2008; @torres2024]. Third, the maintenance of the carbonate framework that provides reef structure [@perry2012]. We aim to assess how these mechanisms interact under projected climate conditions to identify the characteristics of structurally resilient reefs through scenario-based coral population modelling for the Great Barrier Reef ecosystem.

The Great Barrier Reef (GBR) is the world's largest coral reef ecosystem consisting of thousands of individual reefs. They are interconnected through coral and fish larval dispersal patterns mediated by hydrodynamic, spatial, and ecological processes [@cappo2000]. The GBR faces severe projected coral cover declines under climate change [@bozec2025; @bozec2022; @condie2021]. These coral cover declines will be compounded by ocean acidification, tropical cyclones, and crown-of-thorns starfish outbreaks [@death2012; @fox-kemper2021; @seneviratne2021; @sweatman2011; @wolff2018]. The 2023-24 Summer mass bleaching event may be signalling a shift towards this reef future, with some level of bleaching recorded on 79% of surveyed reefs (total 1001 surveyed reefs) [@australianinstituteofmarinescience2024]; @cantin2024\].

@@

We adapted exploratory scenario modelling approach from robust decision making (RDM) approaches for this investigation. In particular, we used parameter sampling and exploratory modelling to investigate a wide range of plausible futures under different assumptions of environmental (climatic) conditions and ecological factors. Exploratory modelling approaches have gained favour in recent years as precise long-term predictions of future conditions are often impossible in complex ecological systems [@conradt2024; @lempert2013]. RDM embraces uncertainty as an inherent characteristic of complex environmental systems [@lempert2010; @lempert2013] and is a core methodology underpinning the development of robust and adaptive management strategies [@haasnoot2013]. Rather than attempting to forecast specific outcomes, RDM aims to systematically explore many possible futures to identify patterns. Allowing for the identification of robust management strategies that perform well across those diverse conditions, as well as conditions in which such strategies may fail. Multiple interacting and compounding sources of uncertainty, and a lack of broad agreement as to their likelihood, create inherently uncertain outcomes. While we did not directly model management decision making in this investigation, we used exploratory modelling to identify the environmental and reef characteristics, such as thermal stress, depth and connectivity, that support favourable outcomes such as recovery and resilience.

@@

To apply exploratory scenario modelling, we employed a framework involving two separate systems: ADRIA (the Adaptive Dynamic Reef Intervention Algorithms platform) and CoralBlox. ADRIA is a multi-criteria decision support platform designed to help reef managers address climate change challenges in coral reef systems [@iwanaga2023]. The ADRIA platform provides functionality for scenario generation and analysis as well as providing the technical infrastructure to couple environmental and ecosystem representations. CoralBlox is a computationally efficient coral population model. It simulates functional/morphological groups of corals over their life cycles, competing for available space over time [@ribeirodealmeida2024]. The coupled framework is referred to as ADRIA-CoralBlox.

@@

Reefs in each time series cluster were compared to identify characteristics that differ between Low, Medium and High clusters. These characteristics include reef median depth, mean time series DHW, ratio of source-to-sink larval connectivity, weighted incoming connectivity, weighted outgoing connectivity and reef carrying capacity (coral habitable area).

@@

Across six of the 29 bioregions reefs in the high cover cluster have lower mean DHW, Log-10 carrying capacity and Log-10 weighted outgoing connectivity values than reefs in other clusters (@sfig-access-esm1-5-bioregion-dhwmean; @sfig-access-esm1-5-bioregion-carrying-capacity; @sfig-access-esm1-5-bioregion-weightedoutgoing). There were no consistent patterns across bioregions in their reefs' Log-10 coral carrying capacity area and cluster assignment (@sfig-access-esm1-5-bioregion-carrying-capacity). Across most bioregions there were not consistent differences between clusters in connectivity metrics such as Log-10 source-to-sink connectivity ratio (@sfig-access-esm1-5-bioregion-logsosi) and Log-10 weighted incoming connectivity (@sfig-access-esm1-5-bioregion-weightedincoming).

@@

@@

Analyses for source-to-sink ratio and weighted incoming and outgoing connectivity metrics suggest that reefs with greater outgoing connectivity and lower incoming connectivity levels have higher probability of being clustered into the Low cluster category compared to other reefs. Additionally, two-way analyses revealed an interaction between reef depth and incoming connectivity in determining cluster category assignment, particularly at the high depth range (@fig-ts-cluster-pdp-twoway-incoming). Two-way analyses revealed that reefs at greater than 10m depth are unlikely to be clustered into the Low cluster category, regardless of their incoming connectivity levels. However, at greater than 10m depth reefs with higher incoming connectivity are more likely to be assigned to a High cluster, while reefs with low incoming connectivity are more likely to be assigned to a Medium cover cluster (@fig-ts-cluster-pdp-twoway-incoming). Two-way analyses, for depth and weighted outgoing connectivity, revealed a slight interaction with the increased low-cluster probability for reefs of high outgoing connectivity primarily concentrated in shallow reefs, while deep reefs with high outgoing connectivity have the greatest probability of being clustered into a high cover cluster (@sfig-ts-cluster-pdp-twoway-outgoing).

![Partial dependence plot indicating the probability of cluster assignment along interacting values of median reef depth (metres) and Log10 weighted incoming connectivity. Colour indicates the probability of assignment to each cluster, with lighter colours indicating higher probability of assignment. Panel A shows the probability of assignment for the Low cover cluster, Panel B shows Medium cover cluster and Panel C shows assignment probability for the High cluster category. Probability values for each cluster are generated by the random forest classifier using each combination of depth and incoming connectivity values along the axes at 50 grid points. The actual distributions of depth and incoming connectivity values for reefs are displayed in panels D and E respectively.](../figs/ts_cluster_rf_twoway_depth_incomingconn_pdp.png){#fig-ts-cluster-pdp-twoway-incoming fig-align="center" width="14.82cm"}

Individual feature analysis for carrying capacity indicates that its influence on cluster assignment is concentrated at the low range of predictor values (@fig-ts-cluster-pdp). This aligns with the right-skewed distribution of carrying capacity across reefs for these three predictor variables. Beyond this low-end range, the partial dependence becomes near constant, indicating that outlier values for these factors are not influential.

Assessing random forest prediction accuracy across bioregions, reveals increasing performance as bioregion mean reef depth increases (@sfig-rf-predictor-perf). This highlights that unexplained variation in the test dataset (33% of variation) may be concentrated in shallow bioregions where the available model predictors do not capture differences between clustered reefs.

@@

Reef median depth had a moderate positive correlation with the number of years a reef maintains coral cover above the assumed carbonate budget threshold required to sustain live coral cover ($\theta_{0}$). Greater median depth was a key characteristic defining reefs that sustain coral cover above $\theta_{0}$ (@fig-access-esm1-5-depth-conn-carbonate; @sfig-access-esm1-5-mean-carb-budget). Unsurprisingly, the number of years above $\theta_0$ substantially decreased as $\theta_0$ is increased, as does the Spearman's correlation. Decreasing correlation levels are likely due to an increasing number of reefs being unable to maintain coral cover above $\theta_0$ as it is increased, with the distribution of years above threshold becoming increasingly right-skewed (@fig-access-esm1-5-depth-conn-carbonate). In contrast, weighted incoming connectivity strength (Log-10 scale) was found to have no correlation with the number of years a reef maintains cover above the required thresholds (@fig-access-esm1-5-depth-conn-carbonate).

![Number of years reef coral cover exceeds the carbonate budget threshold from 2025 to 2075 (y-axis). The assessed carbonate budget thresholds range from 10 to 20% of total reef area. Bracketed x-axis values represent the Spearman's correlation between reef variables (depth - left and Log-10 weighted incoming connectivity - right) and the number of years cover exceeds the carbonate budget threshold. Point colour represents the median reef depth in metres in the left-hand panel, and Log-10 weighted incoming connectivity in the right-hand panel. Red points represent the mean number of years above an assumed carbonate budget threshold across reefs. Deeper reefs are better able to maintain cover above the indicative carbonate budget threshold, while there is no considerable relationship between weighted incoming connectivity and carbonate budget maintenance. Displays results from median time series using ACCESS-ESM1-5.](../figs/ACCESS-ESM1-5/depth_conn_carbonate_comparison.png){#fig-access-esm1-5-depth-conn-carbonate fig-align="center" width="14.82cm"}

@@

Our results show increasing reef depth has a strong positive impact on modelled reef resilience in temporal behaviour and maintenance of positive carbonate budgets under possible future conditions. Reefs with a median depth of 10m or greater, with high incoming connectivity, are the most likely to be assigned to a high cover time series cluster. As an individual feature, incoming connectivity alone does not correlate with maintenance of positive carbonate budgets. These findings highlight the importance of depth and incoming connectivity as coupled selection features when assessing potential reef resilience under future thermal stress.

Reefs in different temporal clusters largely do not differ in mean modelled DHW levels experienced and exhibit overall similar DHW time series. While these reefs experience similar magnitude DHWs based on SST, the proportion of the reefs' coral population impacted by heat stress events depends on reef depth. Reefs at deeper depths are likely to be clustered into High cover clusters. Reefs at shallow depths experience no, or few, years with coral cover exceeding levels likely required to maintain positive carbonate budgets. Movement of shallow reefs towards negative carbonate budgets in the future may contribute to phase shifts in reef dominant CaCO~3~ production. Shifting from coral driven to encrusting algae driven production, or to bioerosion dominated systems with rapid erosion of physical reef frameworks [@dudgeon2010; @perry2008]. While reefs at increased depths may be able to maintain coral driven CaCO~3~ production and accumulation into the future for longer. Our results suggest that within shallow-water reef ecosystems (non-mesophotic, depth \< 30m), depth contributes greatly to reef resilience and the ability of the reef to maintain positive carbonate budgets under future conditions. While similar findings have been demonstrated in the Pacific [@smith2014; @baird2018] and Indo-Pacific [@bridge2014] regions, depth refugia analyses have often focussed on the potential of mesophotic reefs (\> 30m deep) as thermal refugia [@bongaerts2010; @holstein2015; @semmler2017]. In contrast, Venegas et al. [-@venegas2019] found no association between depth and the alleviation of heat stress across a range of shallow reefs in the Pacific.

These findings support the hypothesis that reefs at greater depths will exhibit characteristics of refugia and more prolonged structural resilience under climate change compared to shallower reefs. Additionally, Liu, Callaghan and Baldock [-@liu2024] suggest that rubble instability is inversely correlated with reef depth, possibly limiting juvenile recruitment capacity of shallow reefs. A caveat to these depth-resilience findings is the quality of the bathymetry data used to calculate median reef depth. As detailed previously, estimates of median reef depth are derived from satellite imagery (at $10m^2$ resolution) which can be subject to high degrees of uncertainty, both in terms of the estimated depth at for each pixel as well as their alignment to the established GBRMPA reef outlines. Additionally, we do not incorporate the light-based

Current reef restoration and intervention-investigation efforts on the GBR, such as deployment of coral seeding and fragment devices, target reefs with suitable restoration sites in the 1-10m depth range [@howlett2021; @randall2021; @whitman2024]. While there are no explicit justifications for this depth range, restoration depth is likely constrained by snorkelling and diving deployment and monitoring logistical accessibility, visual identification of sites from the surface, in combination with tourism return value. Our results suggest that of the assessed reefs occurring within the 1-10m depth range (1736 reefs), 57% (991) of reefs fall into the low and medium clusters (across all GCMs at the GBRMPA management region scale), with many unable to maintain positive carbonate budgets under future conditions. Partial dependence analysis revealed a range of depths from 10-15m where the probability of a reef being clustered into the High cover cluster increased substantially. Additionally, two-way partial dependence analysis revealed that high incoming connectivity is important for High cluster assignment for reefs at depths of \>10m. Considering this 10-15m depth range, and targetting well connected reefs, when assessing possible restoration sites may reduce the exposure of planted and seeded corals to high thermal stress, while still maintaining effective logistical access to sites. Managers are also likely to consider the positive potential for flow on effects due to outgoing connectivity when selecting intervention sites [@hock2017]. While reefs with high outgoing connectivity levels were found to have greater probability of being assigned to low cover clusters when assessed individually (@fig-ts-cluster-pdp-twoway-incoming), within the context of reef depth, reefs with high depths and high outgoing connectivity levels had the greatest probability of being clustered into high cover clusters (@sfig-ts-cluster-pdp-twoway-outgoing). These results suggests that interventions targetting deep reefs will also benefit from high incoming and outgoing connectivity levels, allowing for high potential for site resilience and outgoing flow of benefits and recovery. Transient, and difficult to measure, properties such as reef connectivity are likely to increase uncertainty in intervention planning to a greater degree than environmental characteristics such as reef depth. Overall, reef intervention planning is a complex and challenging decision-making process [@anthony2020], and our results suggest that intervention site depth may be an important factor to additionally consider when assessing the long-term resilience of sites.

@@

@@

There are multiple caveats in how coral ecosystem dynamics are modelled in this investigation. We model coral colonies as circular diameter values that are tracked over timesteps through growth. Additionally, we do not model the direct influence of light intensity on coral growth or the negative influence of depth on coral growth rates due to reduced available light levels. [@baker1975; @huston1985; @ruiz-diaz2022]. If included in this investigation, it is possible that reef recovery ability would be reduced at greater depths, reducing the positive effect of depth refugia for temporal resilience. Additionally, reduced growth rates at increased depths may reduce the carbonate budget maintenance of reefs due to lower rates of CaCO~3~ accretion. Due to the availability of quality future projected data we do not consider other important impacts such as waves, cyclones, currents and Crown of Thorns Starfish (CoTS) on coral colony survival and bleaching [@massel1993; @deaker2022].

@@

| Metric | Used in | Definition | Justification |
|------------------|------------------|------------------|------------------|
| Percentage coral cover time series | Timeseries clustering, Carbonate budget analysis | Total coral cover as a percentage of reef area | To summarise coral population health over time |
| Median depth | Time series cluster assessment Carbonate budget analysis | Reef median depth as calculated with zonal raster analysis | Depth is a critical form of environmental refugia from high thermal stress |
| Mean time series DHW | Time series cluster assessment | Reef mean DHW value across time series | Resilient reefs may experience reduced surface DHWs |
| Source - to - sink connectivity ratio | Time series cluster assessment | Ratio of reef outgoing to incoming connectivity | Reefs with greater source than sink connectivity may have lower recovery |
| Weighted incoming connectivity | Time series cluster assessment, Carbonate budget analysis | Incoming connectivity strength weighted by mean time series coral cover of source reefs | High incoming larval flow assists reef recovery |
| Weighted outgoing connectivity | Time series cluster assessment | Outgoing connectivity strength weighted by mean coral cover | High outgoing larval flow may lower recovery capacity of the target reef |
| Carrying capacity | Time series cluster assessment | Reef habitable area \[m^2^\] | Captures the size of the viable reef extent |

: Quantitative metrics used in this investigation

@@

![Partial dependence plot indicating the probability of cluster assignment along interacting values of median reef depth (metres) and Log10 weighted outgoing connectivity. Colour indicates the probability of assignment to each cluster, with lighter colours indicating higher probability of assignment. Panel A shows the probability of assignment for the Low cover cluster, Panel B shows Medium cover cluster and Panel C shows assignment probability for the High cluster category. Probability values for each cluster are generated by the random forest classifier using each combination of depth and weighted outgoing connectivity values along the axes at 50 grid points. The actual distributions of depth and outgoing connectivity values for reefs are displayed in panels D and E respectively.](../figs/ts_cluster_rf_twoway_depth_outgoingconn_pdp.png){#sfig-ts-cluster-pdp-twoway-outgoing fig-align="center" width="14.82cm"}

@@
